treats manual

### 2024-05-15 inst/gitbook/_book/treats_manual.tex
Book up

\title{treats manual}
\author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
\date{2023-12-13}

@@

For more interesting simulations however, you can provide modifiers that actually modify the birth-death process.
We can create one for example that makes species go extinct if their ancestor has a negative trait value.
For that we need to create a \texttt{"modifiers"} object that modifies the \texttt{speciation} process with a specific condition and a specific modification when that condition is met.
For a speciation to occur (and a species to not go extinct), the algorithm draws a random value \emph{x} between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates (\(x < (\lambda \ (\lambda + \mu))\)), else it goes extinct.
\emph{x} is a random value that simulates some stochasticity in the evolutionary simulation.
At every step of the simulation, a different random value \emph{x} is drawn making sure that the resulting speciation or extinction is not predictable.

However, we can create a modifier that introduces less randomness (or a different level of randomness) by forcing the lineage to go extinct every time a specific condition is met, e.g.~if the lineage trait value is negative.
This can be done for example by not randomly generating the value \emph{x} but giving it a fixed value of say 1 so that the estimation of the speciation function \(x < (\lambda \ (\lambda + \mu))\) is always met.
In other words, if the lineage's trait value is negative, make sure that \emph{x} is equal to 1, so that the speciation condition \(x < (\lambda \ (\lambda + \mu))\) is never met.

@@

\begin{verbatim}
## $main
## $main$A
## $main$A$process
## function (x0 = 0, edge.length = 1, Sigma = diag(length(x0)), 
##     ...) 
## {
##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = Sigma * edge.length, 
##         ...)))
## }
## <bytecode: 0x560fcf695180>
## <environment: namespace:treats>
## 
## $main$A$start
## [1] 0
## 
## $main$A$trait_id
## [1] 1
## 
## 
## 
## $background
## NULL
\end{verbatim}

@@

\begin{verbatim}
## Warning in title(...): conversion failure on 'The Ornstein–Uhlenbeck process'
## in 'mbcsToSbcs': dot substituted for <e2>
\end{verbatim}

\begin{verbatim}
## Warning in title(...): conversion failure on 'The Ornstein–Uhlenbeck process'
## in 'mbcsToSbcs': dot substituted for <80>
\end{verbatim}

\begin{verbatim}
## Warning in title(...): conversion failure on 'The Ornstein–Uhlenbeck process'
## in 'mbcsToSbcs': dot substituted for <93>
\end{verbatim}

@@

\begin{verbatim}
## $waiting
## $waiting$fun
## function (bd.params, lineage = NULL, trait.values = NULL, modify.fun = NULL) 
## {
##     return(rexp(1, sum(lineage$n * (bd.params$speciation + bd.params$extinction))))
## }
## <bytecode: 0x560fce0ebbe0>
## <environment: namespace:treats>
## 
## $waiting$internal
## NULL
## 
## 
## $selecting
## $selecting$fun
## function (bd.params, lineage = NULL, trait.values = NULL, modify.fun = NULL) 
## {
##     return(sample(lineage$n, 1))
## }
## <bytecode: 0x560fce0eaf30>
## <environment: namespace:treats>
## 
## $selecting$internal
## NULL
## 
## 
## $speciating
## $speciating$fun
## function (bd.params, lineage = NULL, trait.values = NULL, modify.fun = NULL) 
## {
##     return(runif(1) <= (bd.params$speciation/(bd.params$speciation + 
##         bd.params$extinction)))
## }
## <bytecode: 0x560fce0ea600>
## <environment: namespace:treats>
## 
## $speciating$internal
## NULL
## 
## 
## $call
## $call$waiting
## $call$waiting$fun
## [1] "default"
## 
## 
## $call$selecting
## $call$selecting$fun
## [1] "default"
## 
## 
## $call$speciating
## $call$speciating$fun
## [1] "default"
\end{verbatim}

@@

\begin{longtable}[]{@{}lll@{}}
\toprule
\begin{minipage}[b]{0.20\columnwidth}\raggedright
Function\strut
\end{minipage} & \begin{minipage}[b]{0.28\columnwidth}\raggedright
What it does\strut
\end{minipage} & \begin{minipage}[b]{0.43\columnwidth}\raggedright
Where can it be used\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.20\columnwidth}\raggedright
\texttt{parent.traits}\strut
\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
selects the trait values of the current lineage's parents (i.e.~direct ancestor)\strut
\end{minipage} & \begin{minipage}[t]{0.43\columnwidth}\raggedright
in \texttt{make.modifiers}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright
\texttt{taxa.condition}\strut
\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
provides a trigger for an \texttt{event} dependent on the number of taxa\strut
\end{minipage} & \begin{minipage}[t]{0.43\columnwidth}\raggedright
in \texttt{make.events}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright
\texttt{time.condition}\strut
\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
provides a trigger for an \texttt{event} dependent on time\strut
\end{minipage} & \begin{minipage}[t]{0.43\columnwidth}\raggedright
in \texttt{make.events}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright
\texttt{trait.condition}\strut
\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
provides a trigger for an \texttt{event} dependent on trait values\strut
\end{minipage} & \begin{minipage}[t]{0.43\columnwidth}\raggedright
in \texttt{make.events}\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

\includegraphics[width=6.5in]{../../../../../../tmp/Rtmp6UK3fU/file52c0514edd4f}

@@

\includegraphics[width=6.5in]{../../../../../../tmp/Rtmp6UK3fU/file52c079e156e9}

@@

\includegraphics[width=6.5in]{../../../../../../tmp/Rtmp6UK3fU/file52c02397d39c}

>>>>

\title{treats manual}
\author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
\date{2024-05-15}

@@

For more interesting simulations however, you can provide modifiers that actually modify the birth-death process.
We can create one for example that makes species go extinct if their ancestor has a negative trait value.
For that we need to create a \texttt{"modifiers"} object that modifies the \texttt{speciation} process with a specific condition and a specific modification when that condition is met.
For a speciation to occur (and a species to not go extinct), the algorithm draws a random value \emph{x} between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates (\(x < (\lambda \div (\lambda + \mu))\)), else it goes extinct.
\emph{x} is a random value that simulates some stochasticity in the evolutionary simulation.
At every step of the simulation, a different random value \emph{x} is drawn making sure that the resulting speciation or extinction is not predictable.

However, we can create a modifier that introduces less randomness (or a different level of randomness) by forcing the lineage to go extinct every time a specific condition is met, e.g.~if the lineage trait value is negative.
This can be done for example by not randomly generating the value \emph{x} but giving it a fixed value of say 1 so that the estimation of the speciation function \(x < (\lambda \div (\lambda + \mu))\) is always met.
In other words, if the lineage's trait value is negative, make sure that \emph{x} is equal to 1, so that the speciation condition \(x < (\lambda \div\ (\lambda + \mu))\) is never met.

@@

\begin{verbatim}
## $main
## $main$A
## $main$A$process
## function (x0 = 0, edge.length = 1, Sigma = diag(length(x0)), 
##     ...) 
## {
##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = Sigma * edge.length, 
##         ...)))
## }
## <bytecode: 0x609b29299e58>
## <environment: namespace:treats>
## 
## $main$A$start
## [1] 0
## 
## $main$A$trait_id
## [1] 1
## 
## 
## 
## $background
## NULL
\end{verbatim}

@@

@@

\begin{verbatim}
## $waiting
## $waiting$fun
## function (bd.params, lineage = NULL, trait.values = NULL, modify.fun = NULL) 
## {
##     return(rexp(1, sum(lineage$n * (bd.params$speciation + bd.params$extinction))))
## }
## <bytecode: 0x609b211b6680>
## <environment: namespace:treats>
## 
## $waiting$internal
## NULL
## 
## 
## $selecting
## $selecting$fun
## function (bd.params, lineage = NULL, trait.values = NULL, modify.fun = NULL) 
## {
##     return(sample(lineage$n, 1))
## }
## <bytecode: 0x609b211694e0>
## <environment: namespace:treats>
## 
## $selecting$internal
## NULL
## 
## 
## $speciating
## $speciating$fun
## function (bd.params, lineage = NULL, trait.values = NULL, modify.fun = NULL) 
## {
##     return(runif(1) <= (bd.params$speciation/(bd.params$speciation + 
##         bd.params$extinction)))
## }
## <bytecode: 0x609b21164168>
## <environment: namespace:treats>
## 
## $speciating$internal
## NULL
## 
## 
## $call
## $call$waiting
## $call$waiting$fun
## [1] "default"
## 
## 
## $call$selecting
## $call$selecting$fun
## [1] "default"
## 
## 
## $call$speciating
## $call$speciating$fun
## [1] "default"
\end{verbatim}

@@

\begin{longtable}[]{@{}lll@{}}
\toprule
\begin{minipage}[b]{0.20\columnwidth}\raggedright
Function\strut
\end{minipage} & \begin{minipage}[b]{0.28\columnwidth}\raggedright
What it does\strut
\end{minipage} & \begin{minipage}[b]{0.43\columnwidth}\raggedright
Where can it be used\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.20\columnwidth}\raggedright
\texttt{parent.traits}\strut
\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
selects the trait values of the current lineage's parents (i.e.~direct ancestor)\strut
\end{minipage} & \begin{minipage}[t]{0.43\columnwidth}\raggedright
in \texttt{make.modifiers}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright
\texttt{taxa.condition}\strut
\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
provides a trigger for an \texttt{event} dependent on the number of taxa\strut
\end{minipage} & \begin{minipage}[t]{0.43\columnwidth}\raggedright
in \texttt{make.events}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright
\texttt{age.condition}\strut
\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
provides a trigger for an \texttt{event} dependent on time\strut
\end{minipage} & \begin{minipage}[t]{0.43\columnwidth}\raggedright
in \texttt{make.events}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright
\texttt{trait.condition}\strut
\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
provides a trigger for an \texttt{event} dependent on trait values\strut
\end{minipage} & \begin{minipage}[t]{0.43\columnwidth}\raggedright
in \texttt{make.events}\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

\includegraphics[width=6.5in]{../../../../../../tmp/RtmpcKuqE6/file37eb79e5b154}

@@

\includegraphics[width=6.5in]{../../../../../../tmp/RtmpcKuqE6/file37eb42b1cdb9}

@@

\includegraphics[width=6.5in]{../../../../../../tmp/RtmpcKuqE6/file37eb202f55d4}
